---
title: "Checkpoint 4"
author: "Katya Kelly, Katja McKiernan"
date: "November 17, 2017"
output: html_document
---

### Progress made

Katya - Loaded Google maps using ggmap and plotted the latitudes and longitudes of earthquakes in our data set directly
onto the Google maps. Because the extent of the earthquakes is so large, I created two different base maps: one centered on Point Heiden, Alaska and the other on Los Angeles.

Katja - Worked to add a time_til_quake variable to the dataset. This would give the number of days before/after the nearest major >7 earthquake (both timewise and spacially) that each foreshock/aftershock occurred. So far, I have written half of a function that would create this variable for us without doing it by hand. The function FindMaxMagRow takes in a dataset and returns the row of the earthquake with the largest magnitude within that set. The function CreateTimeToDays will create the time_til_quake variable so it can be attached to the dataset. It compares the time of each quake to the time of the largest quake to get the difference.

```{r}
cleanQuakes = read.csv("cleanQuakes.csv")

library(dplyr)
cleanQuakes1 <- cleanQuakes %>%
  select(-c(X.1))
```

```{r}
library(stringr)
library(lubridate)
#Creating a year, month, and day columns and adding to cleanQuakes1
Year_ <- year(cleanQuakes1$time)
Month_ <- month(cleanQuakes1$time)
Day_ <- day(cleanQuakes1$time)
cleanQuakes1 <- cbind(cleanQuakes1, Year_)
cleanQuakes1 <- cbind(cleanQuakes1, Month_)
cleanQuakes1 <- cbind(cleanQuakes1, Day_)

#lapply(cleanQuakes1$time, as.character)

cleanQuakes1$time <- substring(cleanQuakes1$time, 1, 19)
time_of_Quake <- str_sub(cleanQuakes1$time, -8,-1)
cleanQuakes1 <- cbind(cleanQuakes1, time_of_Quake)

cleanQuakesFin <- cleanQuakes1 %>%
  select(-c(time))
```

```{r}
library(ggmap)
centerLA <- get_map(location = "los angeles",
    color = "color",
    source = "google",
    maptype = "satellite",
    zoom = 4)

centerPH <- get_map(location = "port heiden",
    color = "color",
    source = "google",
    maptype = "satellite",
    zoom = 4)

ggmap(centerLA,
    extent = "device",
    ylab = "Latitude",
    xlab = "Longitude")

ggmap(centerPH,
    extent = "device",
    ylab = "Latitude",
    xlab = "Longitude")


```

```{r}
library(dplyr)
quakesAbove7 <- filter(cleanQuakesFin, mag >= 7)
quakesSince1950 <- filter(cleanQuakesFin, Year_ >= 1950)
quakesAbove7Since1950 <- filter(cleanQuakesFin, Year_ >=1950, mag >= 7)
```


```{r}
library(mapproj)
library(gridExtra)

# color by magnitude, all earthquakes
map1 <- ggmap(centerLA) + geom_point(data=cleanQuakes, aes(x = longitude, y =latitude, color=mag), size = 0.5) +
  scale_color_gradient(low="lightsalmon", high="darkred")
map2 <- ggmap(centerPH) + geom_point(data=cleanQuakes, aes(x=longitude, y=latitude, color=mag), size = 0.5) +
  scale_color_gradient(low="lightsalmon", high="darkred")

# color by magnitude, earthquakes above 7
map3 <- ggmap(centerLA) + geom_point(data=quakesAbove7, aes(x = longitude, y =latitude, color=mag), size = 1) +
  scale_color_gradient(low="lightsalmon", high="darkred")
map4 <- ggmap(centerPH) + geom_point(data=quakesAbove7, aes(x=longitude, y=latitude, color=mag), size = 1) +
  scale_color_gradient(low="lightsalmon", high="darkred")

grid.arrange(map1, map2, map3, map4, ncol=2)
```

```{r}
# color by year, all earthquakes since 1950
map5 <- ggmap(centerLA) + geom_point(data=quakesSince1950, aes(x = longitude, y =latitude, color=Year_), size = 0.5) +
  scale_color_gradient(low="chartreuse", high="chartreuse4")
map6 <- ggmap(centerPH) + geom_point(data=quakesSince1950, aes(x=longitude, y=latitude, color=Year_), size = 0.5) +
  scale_color_gradient(low="chartreuse", high="chartreuse4")

# color by year, earthquakes above 7 since 1950
map7 <- ggmap(centerLA) + geom_point(data=quakesAbove7Since1950, aes(x = longitude, y =latitude, color=Year_), size = 1) +
  scale_color_gradient(low="chartreuse", high="chartreuse4")
map8 <- ggmap(centerPH) + geom_point(data=quakesAbove7Since1950, aes(x=longitude, y=latitude, color=Year_), size = 1) +
  scale_color_gradient(low="chartreuse", high="chartreuse4")

grid.arrange(map5, map6, map7, map8, ncol=2)
```


```{r}
#Create time until big quake or after quake variable - the function Katja's been writing

HIQuake1975 <- filter(cleanQuakes1,  X < 25)
HIQUake1975 <- filter(HIQuake1975, latitude < 50)


FindMaxMagRow <- function(datasub){
  #Finds the row within the input data that has the maximum magnitude
  currentMax = 0
  maxrow = 0
  for(i in 1:nrow(datasub)){
    magnitude = datasub$mag[[i]]
    if(magnitude > currentMax){
      currentMax = magnitude
      maxrow =  i
    }
  }
  return(maxrow)
}



CreateTimeToDays <- function(Qdata){
  #Takes a dataset as an input and adds on a time_til_Quake variable that contains the number of days prior to or after the biggest quake in that dataset; For example if the largest quake occurred on January 5, 1988 at 6 pm, then a quake that occurred in the same region on January 4, 1988 at 5pm would be assigned time_til_Quake =(1.0417). It will be negative if it is an aftershock.
  MaxMagRow = FindMaxMagRow(Qdata)
  QuakeYear = Qdata$Year_[[MaxMagRow]]
  QuakeMonth = Qdata$Month_[[MaxMagRow]]
  QuakeDay = Qdata$Day_[[MaxMagRow]]
  QuakeTime = Qdata$time_of_Quake[[MaxMagRow]]
  QuakeHour = hour(QuakeTime)
  time_til_Quake = rep(0, nrow(Qdata))
  HoursToQuake = 0
  for ( i in 1:nrow(Qdata)){
     if(Qdata$time_of_Quake[[i]] == QuakeTime){
            time_til_Quake[i] = 0
     }
    else{
      if(Qdata$Year_[[i]] == QuakeYear){
        if(Qdata$Month_[[i]] == QuakeMonth){
          if(Qdata$Day_[[i]] == QuakeDay){
            #1. newhour: Get hours from Qdata$Time_of_Quake[[i]]
            time_til_Quake[i] = (QuakeHour-newhour)/24
          }
          else{
            DayDiff = QuakeDay-Qdata$Day_[[i]]
            if(DayDiff==1){
              newhour = 24-(hour(time_of_Quake))
              time_til_Quake[i] = (newhour + QuakeHour)/24
            }
            else{ #if not one but still positive?
              newhour = 24-(hour(time_of_Quake))
              NewDayDiff = DayDiff-1
              newhour = newhour + (NewDayDiff*24)
              time_til_Quake[i] = (newhour + QuakeHour)/24
            }
          }
      }
    }
    }
    
  }
  
}


```

```{r}
summary(cleanQuakesFin)
sum(cleanQuakesFin$mag>=7)
```

### Checkpoint 5 starts here
```{r}
library(ggplot2)
ggplot(data=cleanQuakesFin, aes(x=Year_)) + geom_histogram(color="white")

# make a for loop here to get the counts per year and store them in a vector
# how do i iterate over a column and get the unique values?
sum(cleanQuakesFin$Year_==1925)

#ggplot(data=cleanQuakesFin, aes(x=Year_)) + geom_bar(stat="identity")
```




















